---
title: "hanze_multiomics_course"
author: "Roy Oelen"
date: "2025-06-05"
output: html_document
---

```{r header, include=TRUE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

#!/usr/bin/env Rscript
############################################################################################################################
# Authors: Roy Oelen
# Name: hanze_multiomics_course.Rmd
# Function: show usage of multi-omics data at the single-nucleus level using Seurat/Signac
############################################################################################################################
```

# Hanze multi-omics workshop

This is the multi-omics workshop, where we'll take a look at a multi-omics dataset and use multi-omics data to do cell type annotation


## required software

For this workshop, we'll need R, which you can obtain here: <https://www.r-project.org/> It is also very much recommended to use an Interactive Development Environment, as we'll look at some plots while we are coding. You are free to use any IDE, but I will show examples using Rstudio: <https://posit.co/download/rstudio-desktop/>

We will also need a number of libraries available in R, which we can install using R itself:


```{r install_packages, include=TRUE, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# install packages
install.packages('Seurat')
install.packages('Signac')
install.packages('SeuratDisk')
install.packages('ggplot2')
install.packages('knitr')

```


Which we can next load


```{r load_packages, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# install packages
library('Seurat')
library('Signac')
library('SeuratDisk')
library('ggplot2')
library('knitr')

```


## required data

Additionally, we'll also need some multi-omics data to work with. Data is available here: <https://drive.google.com/drive/folders/1Wtm2vtiwm-CYpENifeNQNfpC-MbW6F9l?usp=sharing>

The only object we'll need for now, is 10k_multimodal_object_processed_nocelltype.rds. This is a multi-omics dataset with both gene expression and chromatin accessibility data for individual immune cells, generated with the 10x multiome protocol

Let's load this data:


```{r load_object, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# set location of where we saved this object
mo_object_loc <- '/groups/umcg-franke-scrna/tmp04/external_datasets/10x_multiome_10k_pbmcs/multimodal/10k_multimodal_object_processed_nocelltype.rds'
# now load this object
mo_object <- readRDS(mo_object_loc)

```


## inspect data

You might not have worked with this type of data before, so let's take a look at it. A Seurat object is one of the ways to store this type of data, and is a rather complex R object. You can just use the variable name here and see there is all sorts of data stored. There is data for a certain amount of cells, at different modalities. We'll start with the expression data first, this is saved under the assays property. Let's see what is there.


```{r check_rna_assay1, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, max.print=10)

# we'll make sure not to print too much
options(max.print=10)

# check the object by itself
mo_object
# check what we have for the assays
mo_object@assays
# and specifically for the RNA one
mo_object@assays$RNA
# which seems to have layers
str(mo_object@assays$RNA@layers) # we'll use str to make sure the print doesn't get too big

```


There are multiple layers there. As often with data analysis, the data needs to be normalized so that it is comparable and you reduce batch effects. These different layers represent different versions of the same data. 'counts' is the raw expression data, 'data' is the data normalized for the sequencing depth, and 'scale.data' is next the same data scaled around zero. We can take a look at the first ten genes of the first ten cells like this


```{r check_rna_assay2, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# check counts
mo_object@assays$RNA@layers$counts[11:20, 11:20]
# and data
mo_object@assays$RNA@layers$data[11:20, 11:20]
# and scale.data
mo_object@assays$RNA@layers$scale.data[11:20, 11:20]

```


What you will see for the first and second layers, are sparse matrices. What this means is that the zeroes in the data are not saved with full precision, but instead what is saved is the number of rows and columns, the locations in the matrix where the value is NOT zero, and what those values are. Expression and accessibility is very zero-inflated, meaning it contains mostly zeroes. By just saving the locations and values of the NOT zero values, we can save a lot memory. The data that is scaled around zero, will not be zero-inflated, and as such will also take up a lot more space in memory.

There are no row and column names here, so this needs a bit of explanation. The rows here are the genes, and the columns are the cells. You usually do not interact with these matrices directly though.

We can also look at the accessibility data, and we see that looks quite similiar, except there are no 'layers', this is because the chromatin library is slightly older. There is also no scale.data.


```{r check_peaks_assay, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# check counts
mo_object@assays$peaks$counts[11:20, 11:20]
# and data
mo_object@assays$peaks$data[11:20, 11:20]

```


Let's look at the data in a different way. We can look at the expression of a specific gene across all cells in a violin plot


```{r gene_exp_all, include=TRUE, eval=TRUE, fig.width = 14, fig.height = 7}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# we are looking at RNA data
DefaultAssay(mo_object) <- 'RNA'
# and making a plot
VlnPlot(mo_object, features = c('CD14', 'FCGR3A'))

```


You might be wondering what everything is in this plot. Every dot is a cell. On the y axis is the expression of that gene for that cell. On the X-axis is the cluster the cell belongs to.


## dimensionality reduction

Expression data alone contains more than 36000 genes, so to profile these cells, we may want to reduce the dimensionality of the data. Usually we run PCA first. We can see how the first two first PCs look


```{r pca_plot, include=TRUE, eval=TRUE, fig.width = 7, fig.height = 7}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a dimension plot of the first two PCs
DimPlot(mo_object, reduction = 'pca', group.by = 'seurat_clusters')

```


Again, each point is an individual cell. Here we already see the data splitting up. What was previously done however, is that KNN-clustering was done on those principal components, to get to the clusters. With PCA we can only look at two dimensions at a time though. We can do further dimensional reduction using UMAP, the result of which we can also visualize


```{r rna_umap_plot, include=TRUE, eval=TRUE, fig.width = 7, fig.height = 7}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of the RNA data
DimPlot(mo_object, reduction = 'umap.rna', group.by = 'seurat_clusters')

```


That was at the RNA level though. If we look at the ATAC data instead, the plot might look a bit different, as the information in the chromatin data is different from the expression data.


```{r atac_umap_plot, include=TRUE, eval=TRUE, fig.width = 7, fig.height = 7}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of the RNA data
DimPlot(mo_object, reduction = 'umap.atac', group.by = 'seurat_clusters')

```


We can however leverage the information of the RNA and the ATAC data together and see how that looks.


```{r wnn_umap_plot, include=TRUE, eval=TRUE, fig.width = 7, fig.height = 7}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of RNA and ATAC data
DimPlot(mo_object, reduction = 'wnn.umap', group.by = 'seurat_clusters')

```


As you see, the clusters overlap quite well with the 2d UMAP. This is part because when the clusters were calculated, this was actually done on the ATAC and RNA data at the same time, instead of only on the RNA or ATAC data.


## cell type annotation (old-fashioned way)

One of the next steps we would usually do, is that we would try to see which exact cell type each cell belongs to. We can plot the expression of specific genes like we did in the Violinplot, but we can also plot the expression of specific genes on top of the UMAP.


```{r wnn_feature_plot, include=TRUE, eval=TRUE, fig.width = 9, fig.height = 9}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of RNA and ATAC data
FeaturePlot(mo_object, reduction = 'wnn.umap', features = c('S100A9', 'CD4', 'CD8A', 'NKG7', 'MS4A1', 'GZMA', 'CD74'))

```


What we would do, is that we would check each cluster and match gene expression levels for those clusters to genes that are marker genes for specific cell types. As you can imagine, this is quite a labour-intensive job, as there are a lot of different marker genes out there. For example this is a relatively low resolution annotation, and it already has a ton of genes to check:


```{r marker_gene_table, include=FALSE, eval=TRUE, fig.width = 9, fig.height = 9}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# make table
marker_df <- data.frame(
  'celltype' = c('monocyte', 
                 'CD4-positive, alpha-beta T cell', 
                 'CD8-positive, alpha-beta T cell',
                 'natural killer cell',
                 'mature B cell', 
                 'mature T cell', 
                 'dendritic cell'), 
  'markers' = c('CTSS, FCN1, NEAT1, LYZ, PSAP, S100A9, AIF1, MNDA, SERPINA1, TYROBP', 
                'IL7R, MAL, LTB, CD4, LDHB, TPT1, TRAC, TMSB10, CD3D, CD3G', 
                'CD8B, CD8A, CD3D, TMSB10, HCST, CD3G, LINC02446, CTSW, CD3E, TRAC', 
                'NKG7, KLRD1, TYROBP, GNLY, FCER1G, PRF1, CD247, KLRF1, CST7, GZMB', 
                'CD79A, RALGPS2, CD79B, MS4A1, BANK1, CD74, TNFRSF13C, HLA-DQA1, IGHM, MEF2C', 
                'CD3D, TRDC, GZMK, KLRB1, NKG7, TRGC2, CST7, LYAR, KLRG1, GZMA', 
                'CD74, HLA-DPA1, HLA-DPB1, HLA-DQA1, CCDC88A, HLA-DRA, HLA-DMA, CST3, HLA-DQB1, HLA-DRB1')
)
# show table
kable(marker_df, caption = 'Marker genes for PBMCs')

```

## cell type annotation (the new way)

Like we have data where there is both chromatin accessibility and gene expression, other multimodal datasets also exist. One technology is CITE-seq, where gene expression and cell surface proteins are measured. These surface marker proteins are the best way to to annotate cells to cell types, as these are much more reliable. One of those datasets is available here: <https://drive.google.com/drive/folders/1oooOROmQU3fi6irALp3YnLoTGbvdV1qw?usp=sharing>

We can next load this object as well


```{r load_reference, include=TRUE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# location of the reference
reference_loc <- '/groups/umcg-franke-scrna/tmp04/external_datasets/azimuth_pbmc_reference/multi.h5seurat'
# read the reference
reference <- LoadH5Seurat(reference_loc)
# update to latest version, as our query is also in this format
reference <- UpdateSeuratObject(reference)

```


Now let's see how this reference looks


```{r check_reference, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, max.print=10)

# we'll make sure not to print too much
options(max.print=10)

# check the object by itself
reference
# check what we have for the assays
reference@assays
# and specifically for the SCT one
reference@assays$SCT
# ADT as well
reference@assays$ADT

```


As this object was created with an older version of the library, you'll see that we are missing the 'layers' sublayer, just as with the chromatin data. That is not super important though. What is more curious is that we don't have an RNA or peaks assay, but instead have an SCT and ADT assay instead. The absence of the peaks assay is not strange, considering I told you this dataset has protein data instead of chromatin data. The ADT is the assay that has the protein data. The SCT assay is here instead of the RNA layer. The SCT assay also is expression data, but that has been normalized in a more specific manner. Again, the details are not super important, but we'll touch upon it in a bit.

We can also check the 2d UMAP in this object


```{r reference_umap_plot, include=TRUE, eval=TRUE, fig.width = 20, fig.height = 10}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of the reference
DimPlot(reference, reduction = 'wnn.umap')

```


As you can see, we are not plotting clusters here, but are going straight for the cell type anotations that were done. They could use protein in addition to RNA for annotations, which you would expect increases their ability to do these annotations.


```{r wnn_reference_feature_plot, include=TRUE, eval=TRUE, fig.width = 9, fig.height = 9}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of RNA and ATAC data
FeaturePlot(reference, reduction = 'wnn.umap', features = rownames(reference@assays$ADT)[11:19])

```


We can use this more confident cell type annotation for our data as well, by projecting our cells on their dataset, and see where in 2d space they land based on their similarity. Then we can transfer the labels from this reference onto the cells in our dataset.

One challenge still, is that our data was normalized different to the reference. That is quickly solved though, by using SCTransform on our data and doing downstream processing to make it comparable to the reference dataset


```{r sct_query, include=TRUE, eval=TRUE, fig.width = 9, fig.height = 9}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# set assay to RNA
DefaultAssay(mo_object) <- 'RNA'
# perform SCTransform
mo_object <- SCTransform(mo_object)
# do PCA based on SCT
mo_object <- RunPCA(mo_object, assay = 'SCT')
# and 2d UMAP
mo_object <- RunUMAP(mo_object, assay = 'SCT', return.model = T, dims = 1:50)

```


Something we have not talked about yet is metadata. We have additional data for each cell that is not directly reflected in their expression/accessiblity/proteins. For example we can have information on from which donor the cells come from. The cell type annotations are usually also stored in the metadata. Let's take a look at the metadata of the reference.


```{r reference_metadata, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# check the reference metadata
head(reference@meta.data)

```


You'll see that there are a number of cell type annotations available. We can actually add all of those to our data. Let's try to finally do this cell type annotation.


```{r reference_mapping, include=TRUE, eval=TRUE, fig.width = 9, fig.height = 9}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# we'll define the function
do_reference_mapping <- function(reference, query){
  # find transfer anchors
  anchors <- FindTransferAnchors(
    reference = reference,
    query = query,
    normalization.method = "SCT",
    reference.reduction = "spca",
    dims = 1:50,
    recompute.residuals = FALSE
  )

  # do the reference mapping
  query <- MapQuery(
    anchorset = anchors,
    query = query,
    reference = reference,
    refdata = list(
      celltype.azi.l1 = "celltype.l1",
      celltype.azi.l2 = "celltype.l2",
      celltype.azi.l3 = "celltype.l3",
      celltype.azi.ADT = "ADT"
    ),
    reference.reduction = "spca",
    reduction.model = "wnn.umap"
  )
  return(query)
}

# and perform the mapping
mo_object <- do_reference_mapping(reference, mo_object)

```


If your computer is not up to snuff with enough memory, you might run out of memory here. Fortunately we can also just load the result


```{r load_ct_object, include=TRUE, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# set location of where we saved this object
mo_object_loc <- '/groups/umcg-franke-scrna/tmp04/external_datasets/10x_multiome_10k_pbmcs/multimodal/10k_multimodal_object_processed.rds'
# now load this object
mo_object <- readRDS(mo_object_loc)

```


We can now look our metadata, and we'll see that we have predicted cell types


```{r mo_metadata, include=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# check the reference metadata
head(mo_object@meta.data)

```


Let's focus on L1 annotation for now, as plot quickly get cluttered with too many cell types. The group.by parameter in DimPlot allows you to plot any categorical value from the metadata onto the 2d UMAP.


```{r wnn_l1_plot, include=TRUE, eval=TRUE, fig.width = 7, fig.height = 7}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of RNA and ATAC data, and set the group.by as the celltype
DimPlot(mo_object, reduction = 'wnn.umap', group.by = 'predicted.celltype.azi.l1')

```


That looks quite nice right? Let's plot that next to our clusters to see how well that overlaps. We'll plot the L2 annotations as well


```{r wnn_l1_cluster_plot, include=TRUE, eval=TRUE, fig.width = 7, fig.height = 7}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# make a plot of the 2d UMAP of RNA and ATAC data, and set the group.by as the celltype
DimPlot(mo_object, reduction = 'wnn.umap', group.by = 'predicted.celltype.azi.l1')
DimPlot(mo_object, reduction = 'wnn.umap', group.by = 'predicted.celltype.azi.l2')
DimPlot(mo_object, reduction = 'wnn.umap', group.by = 'seurat_clusters')

```


we can also check how many cells we have of each cell type. and plot that


```{r cell_type_abundances, include=TRUE, eval=TRUE, fig.width = 5, fig.height = 5}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# get the number of cells for each cell type from the metadata
cell_type_numbers_l1 <- data.frame(table(mo_object@meta.data[['predicted.celltype.azi.l1']]))
# let's update the column names
colnames(cell_type_numbers_l1) <- c('celltype', 'ncells')
# and show that
cell_type_numbers_l1

# we calculate the fraction as well, as that is just the number divided by the total
cell_type_numbers_l1[['fraccells']] <- cell_type_numbers_l1[['ncells']] / sum(cell_type_numbers_l1[['ncells']])

# and we can make some plots
ggplot(data = cell_type_numbers_l1, mapping = aes(x = celltype, y = fraccells, fill = celltype)) +
  geom_bar(stat = 'identity') +
  xlab('Cell type') +
  ylab('Fraction of cells') +
  ggtitle('Fraction of total cells per cell type')

```


There is a lot more that you can do with this type of data. You can compare the expression of genes between samples or cell types, check the cell type proportion difference between samples, compare cases/controls for example, correlate expression and accessiblity, correlate expression and protein abundance etc. For an example, you can see more of this dataset in action right here: github.com/royoelen/process_10x_multiome_10k_pmbc/

